External priors for the next generation of CMB experiments.

### 2015-10-12 Notes/N_eff_prior_paper_v2.tex
new paper text label v2

\date{\today}
\begin{abstract}
The next generation of cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, Dark Matter and Dark Energy. 
Indeed the low level of noise, together with improved angular resolution, will drastically increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure. 
Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments. Here, we examine quantitatively, with a Fisher matrix approach, the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 (S4) experiment.
We find that CMB S4 constraints on neutrino masses can be limited by their degeneracies with the optical depth $\tau$  
if the largest scale at reach is not enough to constraint $\tau$ internally. An external prior on the Hubble constant can improve the CMB S4 constraint significantly while complementary information on $n_{s}$ and the baryon fraction will improve the constraint on $N_{\rm eff}$.
\end{abstract}

@@

\section{Introduction}\label{sec:intro}
Since their early stages, Cosmic Microwave Background (CMB) experiments have been crucial in our understanding of the universe and they will maintain their role also in the near future, thanks to the unprecedented low level of noise and high resolution of their next Stage IV (S4) generation. 
The recent past is indeed reassuring. 
Every new generation of satellite experiments improved the level of sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P,2014A&A...571A..16P,2003ApJS..148..175S,2000ApJ...545L...5H,2000Natur.404..955D}.
This lead us from the observation of the first peak of the CMB temperature power spectrum with COBE to a cosmic variance limited measurement of several peaks with Planck. With the improved sensitivity we extend our understanding of the universe scientific from a proof of its almost flat geometry to a well tested model $\Lambda$CDM with a percent level constraint on its parameters.
The same success characterized ground CMB experiments, where the evolution from DASI \cite{2002ApJ...568...38H} to SPT and ACT \cite{2011ApJ...739...52D} \cite{2011ApJ...743...28K} allowed us to measure the small scale damping tale of the CMB spectrum with increasing accuracy.
These series of successes is far from its end. Indeed the next generation of CMB experiments (S4), now in its planning stage, will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
As it happened in the past, this new sensitivity together with the progress in the measurement of other cosmological probes will improve our understanding of several areas of astrophysics like dark matter, inflation, Dark Energy and neutrinos.

The cosmological dependence to neutrinos is primarily due to their contribution to the number of relativistic species $N_{\rm eff}$ in the early phase of the universe  together with the role they play, given their non zero mass $M_{\nu}$, into the late growth of cosmic structures.
Because relativistic species, like neutrinos, are the main drivers of the cosmic expansion in the early Universe their number affects the expansion rate $H(z)$. This rate can be powerfully tested using the CMB, by carefully comparing the sound horizon scale, obtained from the CMB peaks positions, and the Silk damping scale (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein). 
The future experiments, with their high resolution, will be able to push their measurements deeply into the damping tale of the CMB power spectrum and will unequivocally test the value $N_{\rm eff}=3.046$ predicted by the standard model .
The total mass of neutrinos, on the other hand, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($M_{\nu}<230$ meV from \cite{2014A&A...571A..16P}), neutrinos are still relativistic at the last scattering surface. As other low redshift effects however, massive neutrinos modify the CMB by altering the growth of the large scale structure responsible for the the lensing of the CMB photons. Different neutrino masses consequently lead to different CMB lensing. Stage IV experiment will measure small scales temperature and polarization anisotropies with low noise, drastically improving the lensing reconstruction. This will turn into a  precise constraints of the sum of neutrino masses with a possible hint of the hierarchies of the individual masses. 
Quantitatively, CMB alone will constrain the number of relativistic degree of freedom $N_{\rm eff}$ with a $1\%$ precision and the total mass of neutrinos at $60\%$ with a factor of two improvement on $M_{\nu}$ when priors from baryon acoustic oscillations (BAO) data are introduced \cite{wu:2014}. The future generation of CMB experiments will be a big step in the understanding of the neutrino sector.

As previously mentioned, CMB is also sensitive to the properties of dark energy (see \cite{2010MNRAS.405.2639J}). Thanks to the current generation of experiments we know with extraordinary accuracy its energy density. The challenge for the next generation is to reveal the nature of this mysterious component. For example, a crucial step to rule out part of the proposed models will be to investigate any possible deviations in the equation of state, the ratio of pressure and energy density, from the value $w=-1$ predicted by a cosmological constant. 
Dark energy affects the CMB because it alters the universe's expansion and it consequently changes the distance to the last scattering surface. Furthermore different DE models lead to a different growth of large scale structure which are tested by CMB lensing. Thanks to this sensitivity to cosmological structures present all the way from us to the last scattering surface, CMB will also be a powerful probe of any time dependence of the dark energy equation of state.
However, dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$. Similarly to the neutrino sector, to get competitive DE constraints, CMB experiments will have to rely on external prior coming from BAO and supernova experiments.

The crucial importance of feedback from external experiments is true not only for the dark energy and neutrino sector. 
The future of cosmology will hinge on the ability of combining different probes. In this regard CMB makes no exception: external priors will be fundamental to improve the already tight CMB constraints on cosmological parameters. 
In particular CMB will really benefit from experiments like large scale structure clustering and weak lensing, BAO targeted experiments, and supernovae. 
To improve the synergy between different experiments and to guide the plan of future experiments, in this work we study in detail the dependance of CMB future constrains on the external priors assumed. For all the parameters of the standard $\Lambda$CDM with massive neutrinos and $w\neq-1$ dark energy model we quantify, using a Fisher matrix approach, the CMB constraints as a function of external priors in each of the cosmological parameters.

@@

We want to measure quantitatively the effect of external priors on the  cosmological parameters constraints derived from S4 CMB experiments.
We approach the problem with a Fisher matrix formalism, a simple but powerful technique widely used to forecast future experimental constraints. To apply this technique, we need to specify our cosmological model ($\Lambda$CDM plus neutrinos and dark energy extensions in our case) as well as a set of fiducial parameters and the specifications of S4 CMB experiments.

As usual, we define the fisher matrix elements as the curvature:
\begin{equation}
	\centering
		F_{ij} \equiv - \left\langle\frac{\partial^2 \log \mathcal{L}}{\partial \theta_i \partial \theta_j} \bigg|_{\boldsymbol{\theta} = \boldsymbol{\theta_0}}\right\rangle,
	\label{eqn:Fij_def}
\end{equation}
where $\theta_{i,j}$ represents two of the parameters and $\boldsymbol{\theta_0}$ is the parameters values array that, by definition, maximizes the likelihood. It is important to notice that \refeq{Fij_def} implicitly assume that the likelihood is gaussian. Even if the gaussian approximation has been shown to be appropriate most of the time, some problems have been found in other cases \cite{2012JCAP...09..009W}. Despite this, we do not need the level of accuracy that will require a careful modeling of the likelihoods, because we are looking for the general behavior of parameters errors as a function of external priors.
Moreover the gaussian approximation gets better at smaller scales (high $\ell$ in Fourier space) which, with the exception of the optical depth parameters $\tau$, is where most of the constraining power of the CMB is coming from.

With the usual definition $<a_{\ell m}^{X}a_{\ell' m'}^{Y}>=\delta_{\ell \ell'}\delta_{mm'}C^{X,Y}$, where $a_{\ell m}^{X}$ represent the spherical harmonics coefficients of the field X, the Fisher matrix of a CMB experiment can be rewritten as:
\begin{equation}
 F_{ij} = \sum_\ell \frac{2\ell+1}{2} f_{sky} {\rm Tr} \left(  \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_i} \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_j}  \right).
 \label{eqn:Fij_def2}
 \end{equation}
 In this work we will constrain cosmological parameters with the CMB temperature and E mode polarization together with the reconstructed lensing potential. For this reason, $\boldsymbol{C}_\ell$ in \refeq{Fij_def2} is:
 \begin{eqnarray}
 	\centering
		\mathbf{C}_\ell \equiv \left( \begin{array}{ccc}C_\ell^{TT} + N_\ell^{TT} & C_\ell^{TE} & C_\ell^{Td} \\ C_\ell^{TE} & C_\ell^{EE} + N_\ell^{EE} & 0 \\ C_\ell^{Td} & 0 & C_\ell^{dd} + N_\ell^{dd}\end{array}\right).
	\label{eqn:cov_definition}
\end{eqnarray}
Note that we are neglecting the term $C_\ell^{E\phi}$. As also noticed in previous literature (like \cite{wu:2014,2013PhRvD..87h3008H}) this term contains very little information while adding possible numerical issues.
The term $N_\ell^{X}$ represents the instrumental noise power of the specific experiment and will be discussed in \refssec{cosmo-noise}.
The power of the Fisher approach descend from the Cramer-Rao inequality that relates the error on the parameter $i$, marginalized over all the other parameters, $\sigma_i$, to the Fisher matrix as:
\begin{equation}
\sigma_i \equiv \sigma (\theta_i) = \sqrt{(\mathbf{ F^{-1}})_{ii}}.
\label{eqn:cramer-rao}
\end{equation}
Once $F_{ij}$ is computed following \refeq{Fij_def2} it is straightforwards to get the error $\sigma_i$ from \refeq{cramer-rao}. Furthermore in this context it is easy to introduce external priors on cosmological parameters.
Indeed to introduce priors we simply need to add the Fisher matrixes of the external experiments (before we perform the matrix inversion of \refeq{cramer-rao}), i.e.:
\begin{equation}
F_{\rm total}=F_{\rm CMB}+\sum F_{\rm external}.
\end{equation}
In the same way we can add a prior on a single cosmological parameter just by modyfing the correspondent matrix element.
For example, a $1\%$ prior on $H_{0}$ can be obtained by:
\begin{equation}
F_{H_0 H_0} \rightarrow F_{H_0 H_0} + \frac{1}{(1\% \times H_{0,\text{fid}})^2}.
\end{equation}

We chose a Fisher matrix approach because it is able to forecast future experiments performances without generating mock data, together with the ease of including external priors. 
This technique however introduces also some technical difficulties. 
Indeed it is known \cite{2006astro.ph..9591A} that increasing the number of parameters used in the analysis can lead to numerical issues (see also \cite{2008PhRvD..77d2001V} in the gravitational waves context were several parameters are used).
Fisher matrix indeed can become ill-conditioned: a small change in the fisher matrix led to a big change in its inverse. Because we use \refeq{cramer-rao} this can be a problem for error estimation. 
Even if other methods have been used \cite{2006JCAP...10..013P,2006astro.ph..9591A} Fisher matrices are still the standard method used to forecast future constrains \cite{wu:2014}.
We carefully try to avoid any possible source of errors in computing the elements of \refeq{Fij_def2} and in the matrix inversion of \refeq{cramer-rao}.
We compute the derivatives in \refeq{Fij_def2} using a 5 points formula:
\begin{equation}
\begin{split}
\frac{\partial C}{\partial \theta}\bigg|_{\theta_{0}} \sim & \frac{1}{12 h} [ -C(\theta_{0}+2h)+ 8C(\theta_{0}+h) \\ &-8C(\theta_{0}-h)+C(\theta_{0}-2h)].
\end{split}
\label{eqn:deriv}
\end{equation}

This high order definition allow us to use a bigger gap $h$ around the fiducial parameters $\theta_{0}$. As a consequence the differences of power spectra corresponding to different values are big enough to make possible numerical accuracy issues in computing C negligible.
We also test the robustness of this calculation by changing the gap $h$ in the range $2-7\%$ of the correspondent $\theta_{0}$ without noticing any significant change in the results.
Furthermore we compare our results to similar previous work in the literature \footnote{we thank the authors of \cite{pan:2015} for their collaboration.} obtaining a perfect agreement.
Lastly we implement the same technique of \cite{2006astro.ph..9591A} to avoid possible numerical instability in the marginalizing procedure. This allow us not to invert the entire matrix when we want to marginalize over a set of parameters, in order to minimize numerical issues and conserve parameters degeneracies as much as possible.

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=0.4]{PS_phi_with_noise.pdf}
\caption{Lensing potential power spectrum use in this work has a signal to noise bigger than one up to $\ell\simeq800-1000$.
In the figure: the deflection power spectrum for our fiducial cosmology together with the two examples of lensing reconstruction noise $N^{\phi}$ used in this work.}
\label{fig:phi-cl-noise}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=0.4]{PS_with_noise.pdf}
\caption{The next generation $C^{T,E}$ used in this work are almost cosmic variance limited.
In the figure: CMB power spectrum for our fiducial cosmology together with the instrumental noise used in this work.}
\label{fig:cmb-cl-noise}
\end{center}
\end{figure}

In this subsection we describe the assumptions we made in calculating the elements that goes in \refeq{Fij_def2} and, in particular, the power spectra $C_{\ell}$ and the noise power $N_{\ell}$.
Firstly we parametrize our cosmology using a flat $\nu \Lambda$CDM universe. We allow a set different Dark Energy models by introducing the equations of state parameter $w$ as a varying parameter.
We chose our fiducial parameters following Table 2 of \textit{Planck} best fit \cite{planck-collaboration:2014g}, i.e. $\Omega_c h^2 = 0.12029$, $\Omega_b h^2 = 0.022068$, $A_s = 2.215\times10^{-9}$ at $k_0 = 0.05\ {\rm Mpc}^{-1}$, $n_s = 0.9624$, $\tau = 0.0925$, and $H_0 = 67.11$ km/s/Mpc. Regarding the $\Lambda$CDM extension, we chose a neutrino energy densities $\Omega_{\nu} h^2$=0.0009, which corresponds to $M_{\nu}$ $\simeq$ 85\ meV, a standard $N_{\rm eff}=3.046$ and a dark energy equation of state, $w=-1$.
Given the model, we use CAMB \cite{Lewis:1999bs} to compute the power spectra $C_{\ell}$ at the fiducial values and at those needed to compute derivatives using \refeq{deriv}. Notice that while we vary one parameter in \refeq{deriv} we keep all the others fixed with the exception of $\Omega_{\Lambda}$ which is always changed in order to keep the universe flat ($\Omega_{\rm k} = 0$).

The instrumental noise power $N_{\ell}^{T,E}$ and the lensing reconstruction $N_{\ell}^{\phi}$ in \refeq{Fij_def2} correspond to the expected level for the next generation of CMB experiments (S4).
For the temperature and E-mode polarization of the CMB, together with the improved depth and resolution we also assume that large scale foregrounds, like dust, are under control or negligible. We deal with the presence of point sources poisson noise in the temperature signal by simply discarding all the small scales modes with $\ell>\ell_{\rm T,max}=3000$.
The remaining source of noise, the instrumental noise, is added to the power spectrum in the usual way:
 \begin{equation}
 	\centering
		N^{X}_\ell = s^{\, 2} \exp \left(\ell(\ell+1) \frac{\theta^{\ 2}_{\textsc{fwhm}}}{8\log2}\right),
	\label{eq:beamnoise}
\end{equation}
where $\theta^{\ 2}_{\textsc{fwhm}}$ is the FWHM of the experiment's beam and $s$ represents the instrumental white noise.
We decide to use a level of noise $s = 1.5$ $\mu$K-arcmin for $X=T$ and a beam of $\theta_{\textsc{fwhm}}=1$ arcmin (PRELIMINARY).
Note that $1.5$ $\mu$K-arcmin is the noise in temperature and we need $s \rightarrow s\times \sqrt{2}$ in the case of polarization $ XX' = \{ EE, BB \}$.

Together with E and T we will use the information contained in the lensing potential $\phi$ as it is reconstructed from the CMB. The lensing potential represents the integration along the line of site of the gravitational potential and it leaves its signature in the CMB, both in temperature and polarization, by bending the trajectory of CMB photons. This introduces non gaussianities that couple different, otherwise independent, CMB modes and it can be reconstructed using a quadratic estimator technique \cite{okamoto:2003,hu:2002}.
For the noise $N_\ell^{\phi\phi}$ associated to the reconstructed $\phi$ we follow the non-iterative technique of \cite{okamoto:2003,hu:2002}.

%\begin{eqnarray}
%\centering
%	s\ [\;\mu {\rm K.arcmin}\;] \equiv \frac{ {\rm NET}\ [\; \mu{\rm K.}\sqrt{s}\;] \times \sqrt{f_{sky} \ [\;{\rm arcmin}^2\;] }}{ \sqrt {N_{\rm det} \times Y \times \Delta T\ [\;{\rm s }\;]}}.
%	\label{eq:sensitivity_definition}
%\end{eqnarray}
%

%\begin{eqnarray}
%	T_{\nu} = \left( \frac{4}{11} \right)^{1/3} T_{\gamma} 
%	\label{eq:tnu_propto_tgamma}
%\end{eqnarray}

\section{Results \label{sec:results}}
In this section we present our main results. We will start from..

@@

Small scale structure formation is suppressed if fast-moving neutrinos comprise a significant part of the matter budget. This effect, probed by CMB lensing, will allow us to determine how much of the matter density is made up of neutrinos. Since the active neutrino number densities are known in the standard model, this can be directly transformed into a constraint on the the sum of the neutrino masses, $\sum m_\nu$. This is particularly exciting because there is a lower limit on $\sum m_\nu>0.05$ eV that emerges from oscillation experiments.

\begin{figure}[htbp]
\begin{center}
\includegraphics{prior_omnuh2_snow_mass_lmin=4_lmax=4499.pdf}
\caption{Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters, as in \reffig{prior_massless_neutrinos}.  Note that constraints would be improved significantly  by distance measurements that constrain $\Omega_ch^2$. The lower limit on the sum of the neutrino masses is 50 meV.}
\label{fig:prior_omeganuh2}
\end{center}
\end{figure}
%==========

>>>>

\date{\today}
\begin{abstract}
The next generation of cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, Dark Matter and Dark Energy. 
Indeed the low level of noise, together with improved angular resolution, will allow a precise measurement of the CMB polarization anisotropies as well as an improved reconstruction of the lensing potential of high redshift large scale structure. 
Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments. Here, we examine quantitatively, with a Fisher matrix approach, the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 (S4) experiment.
We find that CMB S4 constraints on neutrino masses can be limited by their degeneracies with the optical depth $\tau$ unless the largest scale at reach for CMB S4 will be large enough to constraint $\tau$ internally thought the reionization bump. An external prior on the Hubble constant can improve the CMB S4 constraint significantly while complementary information on $n_{s}$ and the baryon fraction will improve the constraint on $N_{\rm eff}$.
\end{abstract}

@@

\section{Introduction}\label{sec:intro}
Since their early stages, Cosmic Microwave Background (CMB) experiments have been crucial in our understanding of the universe. They have the potential to maintain this important role in the near future thanks to the unprecedented low level of noise and high resolution of their next Stage IV (S4) generation. 
Indeed, the next generation of CMB experiments (S4), now in its planning stage, will measure the E-mode polarization with almost cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction.
This new sensitivity, in particular if supported with the essential synergy with other cosmological probes, will be a crucial step in our understanding of dark matter, inflation, Dark Energy and neutrinos.

Neutrinos properties can be effectively constrained with CMB measurements because they influence both the early history of the universe and the later stage of structure formation. 
Because relativistic species, like neutrinos, are the main drivers of the cosmic expansion in the early Universe their effectivenumber $N_{\rm eff}$ affects the expansion rate. This can be powerfully tested using the CMB, by comparing the sound horizon scale, obtained from the CMB peaks positions, and the Silk damping scale (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein). 
The future experiments, with their high resolution, will be able to measure the CMB power spectrum, at small scales, deeply into the damping tale, testing any possible deviations from $N_{\rm eff}=3.046$, the value predicted by the standard model.
The total mass of neutrinos, on the other hand, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($M_{\nu}<230$ meV from \cite{2014A&A...571A..16P}), neutrinos are still relativistic at the last scattering surface. However massive neutrinos influence the growth of structure at scales shorter than their free streaming. Different neutrino masses consequently lead to different CMB lensing. Stage IV experiment will measure small scales temperature and polarization anisotropies with low noise, drastically improving the lensing reconstruction. This will turn into such a precise constraints of the sum of neutrino masses to get a possible hint of the hierarchies of the individual masses. 
Quantitatively, CMB alone will constrain the number of relativistic degree of freedom $N_{\rm eff}$ with a $1\%$ precision and the total mass of neutrinos at $45\%$ with a factor of two improvement on $M_{\nu}$ when priors from baryon acoustic oscillations (BAO) data are introduced \cite{wu:2014}. The future generation of CMB experiments will be a big step in the understanding of the neutrino sector.

As previously mentioned, CMB is also a powerful probe of the dark energy properties (see \cite{2010MNRAS.405.2639J}). The challenge for future experiments is to go beyond the now well measured energy density and to determine other dark energy properties.
Among these, a crucial step would be to constraint the equation of state, the ratio of pressure and energy density, looking for possible deviations from the value $w=-1$ expected in a cosmological constant scenario. 
CMB is sensitive to Dark energy because this component alters the universe's expansion and it consequently changes the distance to the last scattering surface. Furthermore different DE models lead to a different growth of large scale structure which are tested by CMB lensing.
However, dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$. As a consequence, similarly to the neutrino sector, to get competitive DE constraints, CMB experiments will have to rely on external prior coming from BAO and supernova experiments.

The crucial importance of the feedback from external experiments is true not only for the dark energy and the neutrino sector. 
In the future, cosmology will hinge on the ability of combining different probes. In this regard CMB makes no exception: external priors will be fundamental to improve the already tight CMB constraints on cosmological parameters. 
In particular CMB will really benefit from experiments like large scale structure clustering and weak lensing, BAO targeted experiments, and supernovae. 
To improve the synergy between different experiments and to guide the plan of future experiments, in this work we study the dependance of CMB future constrains on the external priors assumed. For all the parameters of the standard $\Lambda$CDM plus massive neutrinos ($\Omega_c h^2,\Omega_b h^2,A_s,n_s,\tau,H_0,M_{\nu}$) together with possibly testable non-standard extension ($N_{\rm eff},w,w_{a}$) we quantify, using a Fisher matrix approach, the CMB constraints as a function of external priors in each of the cosmological parameters.

@@

To measure quantitatively the impact of external priors on the CMB ability to constrain cosmological parameters we use a Fisher matrix formalism. In this section we will quickly review  the technique and we will then present our fiducial cosmological model together with the experimental specifications of S4 CMB we used.

As usual, we define the Fisher matrix elements as the curvature of the likelihood:
\begin{equation}
	\centering
		F_{ij} \equiv - \left\langle\frac{\partial^2 \log \mathcal{L}}{\partial \theta_i \partial \theta_j} \bigg|_{\boldsymbol{\theta} = \boldsymbol{\theta_0}}\right\rangle,
	\label{eqn:Fij_def}
\end{equation}
where $\theta_{i,j}$ represents two of the cosmological parameters and $\boldsymbol{\theta_0}$ is the fiducial values array that, by definition, maximizes the likelihood. It is important to notice that \refeq{Fij_def} implicitly assume that the likelihood is gaussian. Even if the gaussian approximation has been shown to be appropriate most of the time, exceptions exist in the literature \cite{2012JCAP...09..009W}. However, the level of accuracy needed in this work does not require a more careful modeling of the likelihoods. Moreover the gaussian approximation gets better at smaller scales (high $\ell$ in Fourier space) which, with the exception of the optical depth parameters $\tau$, is where most of the constraining power of the CMB is coming from.

For CMB experiments the Fisher matrix can be related to the power spectrum $\boldsymbol{C}_\ell$ by:
\begin{equation}
 F_{ij} = \sum_\ell \frac{2\ell+1}{2} f_{\rm sky} {\rm Tr} \left(  \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_i} \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_j}  \right).
 \label{eqn:Fij_def2}
 \end{equation}
In this work we constrain cosmological parameters with CMB temperature and E mode polarization together with the reconstructed lensing potential of large scale structure. For this reason, $\boldsymbol{C}_\ell$ in \refeq{Fij_def2} is:
 \begin{eqnarray}
 	\centering
		\mathbf{C}_\ell \equiv \left( \begin{array}{ccc}C_\ell^{TT} + N_\ell^{TT} & C_\ell^{TE} & C_\ell^{T\phi} \\ C_\ell^{TE} & C_\ell^{EE} + N_\ell^{EE} & 0 \\ C_\ell^{T\phi} & 0 & C_\ell^{\phi} + N_\ell^{\phi}\end{array}\right).
	\label{eqn:cov_definition}
\end{eqnarray}
Note that we are neglecting the term $C_\ell^{E\phi}$. As also noticed in previous literature \cite{wu:2014,2013PhRvD..87h3008H} this term contains very little information while adding possible numerical issues.
The terms $N_\ell^{X}$ represents the instrumental noise power of the specific experiment and will be described at the end of this section.

A lower bound on the error on the parameter $i$, marginalized over all the other parameters, $\sigma_i$, can be easily obtained as:
\begin{equation}
\sigma_i \equiv \sigma (\theta_i) \geq \sqrt{(\mathbf{ F^{-1}})_{ii}}.
\label{eqn:cramer-rao}
\end{equation}
Furthermore it is easy to introduce external priors on cosmological parameters.
Indeed we simply need to add the Fisher matrixes of the external experiments (before we perform the matrix inversion of \refeq{cramer-rao}), i.e.:
\begin{equation}
F_{\rm total}=F_{\rm CMB}+\sum F_{\rm external}.
\end{equation}
In the same way we can add a prior on a single cosmological parameter just by modyfing the correspondent matrix element.
For example, a $1\%$ prior on $H_{0}$ can be added by:
\begin{equation}
F_{H_0 H_0} \rightarrow F_{H_0 H_0} + \frac{1}{(1\% \times H_{0,\text{fid}})^2}.
\end{equation}

We chose a Fisher matrix approach because it is able to forecast future experiments performances without generating mock data and easily including external priors. 
This technique however introduces also technical difficulties. 
Indeed it is known \cite{2006astro.ph..9591A} that increasing the number of parameters used in the analysis can lead to numerical issues (see also \cite{2008PhRvD..77d2001V} in the gravitational waves context were several parameters are used).
Fisher matrix indeed can become ill-conditioned: a small change in the fisher matrix lead to a big change in its inverse. Because we use \refeq{cramer-rao} this can be a problem for error estimation. 
Even if other methods have been used \cite{2006JCAP...10..013P,2006astro.ph..9591A} Fisher matrices are still the standard procedure used to forecast future constrains \cite{wu:2014}.
We carefully try to avoid any possible source of errors in computing the elements of \refeq{Fij_def2} and in the matrix inversion of \refeq{cramer-rao}.
We compute the derivatives in \refeq{Fij_def2} using a 5 points finite difference formula:
%\begin{equation}
%\begin{split}
%\frac{\partial C}{\partial \theta}\bigg|_{\theta_{0}} \sim & \frac{1}{12 h} [ -C(\theta_{0}+2h)+ 8C(\theta_{0}+h) \\ &-8C(\theta_{0}-h)+C(\theta_{0}-2h)].
%\end{split}
%\label{eqn:deriv}
%\end{equation}
this high order approach allows us to use bigger steps around the fiducial parameters to compute derivatives. As a consequence the differences of power spectra corresponding to different values are big enough to make possible numerical accuracy issues in computing $\boldsymbol{C}_\ell$ negligible.
We also test the robustness of this calculation by changing the derivative steps in the range $2-7\%$ of the correspondent $\theta_{0}$ without noticing any significant change in the results.
%Furthermore we compare our results to similar previous work in the literature \footnote{we thank the authors of \cite{pan:2015} for their collaboration.} obtaining a perfect agreement.
%Lastly we implement the same technique of \cite{2006astro.ph..9591A} to avoid possible numerical instability in the marginalizing procedure. This allow us not to invert the entire matrix when we want to marginalize over a set of parameters, in order to minimize numerical issues and conserve parameters degeneracies as much as possible.

%\begin{figure}[htbp]
%\begin{center}
%\includegraphics[scale=0.4]{PS_phi_with_noise.pdf}
%\caption{Lensing potential power spectrum use in this work has a signal to noise bigger than one up to $\ell\simeq800-1000$.
%In the figure: the deflection power spectrum for our fiducial cosmology together with the two examples of lensing reconstruction noise $N^{\phi}$ used in this work.}
%\label{fig:phi-cl-noise}
%\end{center}
%\end{figure}
%
%\begin{figure}[htbp]
%\begin{center}
%\includegraphics[scale=0.4]{PS_with_noise.pdf}
%\caption{The next generation $C^{T,E}$ used in this work are almost cosmic variance limited.
%In the figure: CMB power spectrum for our fiducial cosmology together with the instrumental noise used in this work.}
%\label{fig:cmb-cl-noise}
%\end{center}
%\end{figure}

We compute the power specctra $C_{\ell}$ and the noise power $N_{\ell}$ in \refeq{Fij_def2} using CAMB.
Our base cosmology is a flat $\nu \Lambda$CDM. 
We then extend the parameter space by introducing $N_{\rm eff}$ and $w,w_{a}$ as free parameters.
We chose our fiducial parameters following Table 2 of \textit{Planck} best fit \cite{planck-collaboration:2014g}, i.e. $\Omega_c h^2 = 0.12029$, $\Omega_b h^2 = 0.022068$, $A_s = 2.215\times10^{-9}$ at $k_0 = 0.05\ {\rm Mpc}^{-1}$, $n_s = 0.9624$, $\tau = 0.0925$, $H_0 = 67.11$ km/s/Mpc and a neutrino total mass $M_{\nu}$ $\simeq$ 85\ meV. 
Regarding the $\Lambda$CDM extensions, we chose, a standard $N_{\rm eff}=3.046$ and a dark energy equation of state, $w=-1,w_{a}=0$.
It is important to notice that while we vary one parameter to compute derivatives in \refeq{Fij_def2} we keep all the others fixed with the exception of $\Omega_{\Lambda}$ which is always changed in order to keep the universe flat ($\Omega_{\rm k} = 0$).

The instrumental noise power $N_{\ell}^{T,E}$ and the lensing reconstruction $N_{\ell}^{\phi}$ in \refeq{Fij_def2} correspond to the expected level for the next generation of CMB experiments (S4).
For the temperature and E-mode polarization of the CMB, together with the improved depth and resolution we also assume that large scale foregrounds, like dust, are under control or negligible. This allow us to use all the polarization power spectrum multipoles all the way up to $\ell_{\rm E,max}=5000$. We deal with the presence of point sources poisson noise in the temperature signal by simply discarding all the small scales modes with $\ell>\ell_{\rm T,max}=3000$.
The remaining source of noise, the instrumental noise, is added to the power spectrum in the usual way:
 \begin{equation}
 	\centering
		N^{X}_\ell = s^{\, 2} \exp \left(\ell(\ell+1) \frac{\theta^{\ 2}_{\textsc{fwhm}}}{8\log2}\right),
	\label{eq:beamnoise}
\end{equation}
where $\theta^{\ 2}_{\textsc{fwhm}}$ is the FWHM of the experiment's beam and $s$ represents the instrumental white noise.
We decide to use a level of noise $s = 1.5$ $\mu$K-arcmin for $X=T$ and a beam of $\theta_{\textsc{fwhm}}=1$ arcmin (PRELIMINARY).
Note that $1.5$ $\mu$K-arcmin is the noise in temperature and we assume that $s \rightarrow s\times \sqrt{2}$ in the case of polarization $ XX' = \{ EE, BB \}$.

%Together with E and T we will use the information contained in the lensing potential $\phi$ as it is reconstructed from the CMB. The lensing potential represents the integration along the line of site of the gravitational potential and it leaves its signature in the CMB, both in temperature and polarization, by bending the trajectory of CMB photons. This introduces non gaussianities that couple different, otherwise independent, CMB modes and it can be reconstructed using a quadratic estimator technique \cite{okamoto:2003,hu:2002}.
The noise $N_\ell^{\phi\phi}$ associated to the reconstructed $\phi$ is modeled assuming an iterative reconstruction technique. Indeed it has been shown \cite{seljak:2004} that the low level of noise expected in CMB S4 experiment will be in the regime  where this technique improve our constraints compared to the quadratic estimator introduced in \cite{okamoto:2003,hu:2002}.

\section{Results \label{sec:results}}
In this section we present our main results. We will start from our base $\nu \Lambda$CDM and we then explore possible extensions.

Small scale structure formation is suppressed if fast-moving neutrinos comprise a significant part of the matter budget. This effect, probed by CMB lensing, will allow us to determine the contribution of neutrinos to the total matter density. Since the active neutrino number densities are known in the standard model, this can be directly transformed into a constraint on the the sum of the neutrino masses, $\sum m_\nu$. This is particularly exciting because we now know that neutrinos are massive with a lower limit on $\sum m_\nu>0.05$ eV that emerges from oscillation experiments.

\begin{figure}[htbp]
\begin{center}
\includegraphics{prior_omnuh2_snow_mass_lmin=4_lmax=4499.pdf}
\caption{Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters, as in \reffig{prior_massless_neutrinos}.  Note that constraints would be improved significantly  by distance measurements that constrain $\Omega_ch^2$. The lower limit on the sum of the neutrino masses is 50 meV.}
\label{fig:prior_omeganuh2}
\end{center}
\end{figure}
%==========
%==========

\begin{figure}[htbp]
\begin{center}
\includegraphics{prior_omnuh2_snow_mass_lmin=50_lmax=3000_ndet=10000_fsky=0.75}
\caption{Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters, as in \reffig{prior_massless_neutrinos}.  Note that constraints would be improved significantly  by distance measurements that constrain $\Omega_ch^2$. The lower limit on the sum of the neutrino masses is 50 meV.}
\label{fig:prior_omeganuh2_tau}
\end{center}
\end{figure}
%==========

\reffig{prior_omeganuh2} shows that, without any external priors, upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 30 meV, corresponding to close to a 2-sigma detection even in the worst case scenario. Adding in external priors helps this significantly though, as discussed in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}. There, the prior was deescribed as ``DESI BAO'', that is a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by DESI. Here, the same improvement is parametrized by a strong prior on $\Omega_ch^2$, which indeed determines these low-redshift distances. As found in \cite{2013arXiv1309.5383A}, the projected error falls below 20 meV with this prior. Adding in other priors would help further.

@@
